---
title: GSEA analysis
author: Carl Murie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document

---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(GSEABase)
library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(limma)
library(DT)
library(magrittr)
```

```{r read}
fdr_cut <- 0.25
min_gene_set <- 5

vahey_eset <- readRDS(here("data", "gpl_eset.Rds"))
```

```{r gsea_setup}
## set up GSEA analysis
btm_gtm <- getGmt(here("data/h.all.v5.0.symbols.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(vahey_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```

#Analysis

The hallmark gene-sets from the Broad Institute's Molecular Signatures Database were used for the GSEA analysis.

##GSEA: visit T4 protected vs non-protected (Protected - NonProtected)

A FDR cutoff of <= `r fdr_cut` was used.  
  
```{r gsea1}

## filter data for relevant time points and disease status

pData(vahey_eset) %>% 
  filter(visit == "T4") %>%
  filter(disease == "Protected" | disease == "NonProtected") %$%
    geo_id ->
  sample_index

gsea_eset <- vahey_eset[,sample_index]
pData(gsea_eset) %>% 
  mutate(disease=factor(disease, levels=c("NonProtected", "Protected"))) ->
  pData(gsea_eset)
```

###Sample sizes  

```{r samp_deg}
tab <- countSampleSizes(pData(gsea_eset), "visit", "disease")
headerKable(tab, labels="disease", levels=2)
```

```{r voom}

design1 <- model.matrix(~disease, data=pData(gsea_eset))
v1 <- voom(exprs(gsea_eset), design=design1)
btm_camera <- as_tibble(camera(v1, btm_ind, design1, 
                               contrast="diseaseProtected"), rownames="geneset")
num_sig <- sum(btm_camera$FDR <= fdr_cut)
```

There were `r num_sig` significant gene-sets based on fdr cutoff.  
  
```{r gsea_output}
btm_camera %>% 
  filter(FDR < fdr_cut) %>% 
  datatable() %>% 
  formatRound(columns = c("PValue", "FDR"), digits = 4)
```
