---
title: "Monocyte Lymphocyte ratio prediction for disease"
subtitle: "CiberSort"
author: "Carl Murie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(DT)
library(magrittr)
library(survival)   ## clogit
```

```{r data}
vah_eset <- readRDS(here("data/gpl_eset.Rds"))
ciber <- read_csv(here("data/vahey_ciber_cells.csv"))
```

#Test monocyte/lymphocyte (ML) ratios for disease prediction  
  
```{r ciber}

## extract lymphocyte cell names
lymph_names <- colnames(ciber)[str_detect(colnames(ciber), 
                                          "T cells|B cells|NK cells|Plasma cells")]

## extract non-lymphocyte cell names
not_lymph_names <- colnames(ciber)[!str_detect(colnames(ciber), 
                                          "T cells|B cells|NK cells|Plasma cells")]
not_lymph_names <- not_lymph_names[-c(1:2,12:14)]
not_lymph_names <- c(not_lymph_names, c("", "", ""))

## calculate monocyte/lymphcyte ratios
ciber %<>%
  mutate(Lymphocytes= apply(ciber[,lymph_names], 1, sum)) %<>%
  mutate(ml_ratio=Monocytes/Lymphocytes)

## join with data
pData(vah_eset) %>% 
  filter(disease=="Protected" | disease == "NonProtected") %>%
  select(-c("description", "parasitemia", "pid", "chip")) %>%
  dplyr::left_join(ciber[,c("Input Sample", "ml_ratio", "Monocytes")], 
                   by=c("geo_id"="Input Sample")) ->
  dat

```

The CiberSort deconvolution method was applied to the Vahey microarray data to produce monocyte/lymphocyte ratios. CiberSort produces an estimate of Monocyte proportion and a number of Lymphocyte cell subset proportions.  
  
The following lymphocyte subsets were summed to calculate the lymphocyte proportion:  
  
```{r names}
outie <- cbind(lymph_names, not_lymph_names)
colnames(outie) <- c("Lymphocytes", "not Lymphocytes")
headerKable(outie)
```
  
The monocyte/lymphocyte ratio was calculated by dividing the monocyte proportion by the lymphocyte proportion.  
  
##Sample Sizes  
  
```{r samples} 
countSampleSizes(dat, "visit", "disease") %>%
  headerKable(labels="disease", levels=2)
```

##Plot ML ratios against disease  
  
```{r plot}
ggplot(dat, aes(x=disease, y=ml_ratio, color=disease)) + geom_boxplot() + 
  facet_wrap(~visit)
```

##Test ML ratios for disease prediction  
  
clogit regression will be applied with the following formula:  
  
disease ~ ml_ratio   
  
```{r clogit}

## convert disease to binary clogit variable
dat %<>%
 mutate(dis_logit:=as.numeric(plyr::mapvalues(as.character(disease),
                                          from=c("Protected","NonProtected"), 
                                          to=c(0,1)))) %<>%
  data.table()

## execute test
logit_pvals <- dat[,summary(clogit(dis_logit~ml_ratio))$coefficients["ml_ratio", 5], by=visit]

colnames(logit_pvals)[2] <- "Pvalue"
logit_pvals %>%
  headerKable()
```
  
  
##Plot Monocyte populationss against disease  
  
```{r plot_mono}
ggplot(dat, aes(x=disease, y=Monocytes, color=disease)) + 
  geom_boxplot() + xlab("disease") + facet_wrap(~visit)
```

