---
title: Experimental Design
author: Carl Murie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document

---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(GEOquery)
library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(limma)
```



```{r read}
vahey_eset <- readRDS(here("data", "gpl_eset.Rds"))

melt(exprs(vahey_eset), id.vars=1:ncol(exprs(vahey_eset))) %>%
  rename(Gene=Var1, 
         geo_id=Var2, 
         read=value) %>%
  mutate(Gene=as.character(Gene),
         geo_id=as.character(geo_id)) %>%
  inner_join(pData(vahey_eset)[,c("geo_id", 
                                  "pid", 
                                  "parasitemia", 
                                  "disease", 
                                  "visit",
                                  "chip")], 
             by="geo_id") %>%
  mutate(visit = factor(visit, levels=c("T0", "T1", "T2", "T3", "T4", "T5")),
                        ordered=TRUE) ->
  vahey_dt

 num_controls <- sum(vahey_eset$parasitemia=="InfectivityControl")
```

#Experimental Design

Microarray gene expression assay applied to PBMC samples of subjects exposed to RTS,S/AS01 or RTS,S/AS02 vaccine.  
  
**Microarrays**  
HGU133A 2.0 - 13515 genes (used for all non-T5 samples)  
HGU133 Plus 2.0 - 23521 genes (used for all T5 samples)  
  
The genes in the HGU133A are a subset of the genes in the HG133 Plus chip. The intersection of the genes from the 2 chips were used in this analysis which resulted in 13515 genes.  
  
**Factors**:  
  
* **visit**
    + T0 - study entry (baseline)
    + T1 - day of third vaccination
    + T2 - 24h after T1
    + T3 - 72h after T1
    + T4 - 2 weeks after T1
    + T5 - 5 days after challenge
* **parasitemia** - response to malaria challenge
    + Protected - vaccinated and protected from challenge  
    + Delay - vaccinated and delayed parasitemia  
    + NonProtected - vaccinated and not protected from challenge  
    + InfectivityControl - not vaccinated and challenged  
    + Unknown - unknown
* **disease** - identical to parasitemia but Delay was changed to NonProtected  

##Sample sizes  
  
There are `r num_controls` samples that are labeled InfectivityControl. They do not have time-points
  
```{r samples}

tab <- countSampleSizes(pData(vahey_eset), "visit", c("parasitemia"))
headerKable(tab, labels="parasitemia", levels=5)
```
