---
title: GSEA analysis - day of vaccination (T1) vs all other later time periods
subtitle: time effect
author: Carl Murie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(GSEABase)
library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(limma)
library(DT)
library(magrittr)
library(heatmap3)
```

```{r read}
fdr_cut <- 0.2
min_gene_set <- 5
form <- "~visit"        ## formula used in camera

vahey_eset <- readRDS(here("data", "gpl_eset.Rds"))
```

```{r gsea_setup}
## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(vahey_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```

# GSEA Analysis

The Blood Transcript Modules (BTM) from Shuzhao et al. Molecular signatures of antibody responses derived from a systems biological study of 5 human vaccines, 2014 were used for the GSEA analysis.
  
## Day of vaccine (T1) vs visits T2 to T5: time effect  
  
* **visit**
    + T0 - study entry (baseline)
    + T1 - day of third vaccination
    + T2 - 24h after T1
    + T3 - 72h after T1
    + T4 - 2 weeks after T1
    + T5 - 5 days after challenge  
  
A FDR cutoff of <= `r fdr_cut` was used.  
  
The formula used in the camera function is: `r form`  
  
```{r gsea1}

## filter data for relevant time points and disease status

pData(vahey_eset) %>% 
  filter(!is.na(visit)) %>%           ## remove controls
  filter(disease != "Unknown") %$%    ## remove Unknown subjects - no samples for T0 or T5
    geo_id ->
  sample_index

gsea_eset <- vahey_eset[,sample_index]
```

### Sample size  
  
```{r sample_size}
tab <- countSampleSizes(pData(gsea_eset), "visit", "disease")
headerKable(tab, labels="disease", levels=2)
```
  
## Comparisons {.tabset .tabset-fade .tabset-pills}  
  
### T1 versus T2 (T2 - T1)  
  

```{r T1vsT2}

## get subset
pData(gsea_eset) %>% 
  filter(visit == "T1" | visit == "T2") %$%         
    geo_id ->
  visit_index

gsea_sub <- gsea_eset[,visit_index]

pData(gsea_sub) %>%
  mutate(visit=factor(visit, levels=c("T1", "T2"))) ->
  pData(gsea_sub)

design1 <- model.matrix(formula(form), data=pData(gsea_sub))
v1 <- voom(exprs(gsea_sub), design=design1)
btm_camera <- as_tibble(camera(v1, btm_ind, design1, 
                               contrast="visitT2"), rownames="geneset")
num_sig <- sum(btm_camera$FDR <= fdr_cut)

btm_camera %>% 
  mutate(time = "T2") ->
all_res


btm_camera %>% 
  filter(FDR < fdr_cut) %>% 
  headerKable() %>%
  kableExtra::scroll_box(height="300px")
```

### T1 versus T3 (T3 - T1)  
  

```{r T1vsT3}

## get subset
pData(gsea_eset) %>% 
  filter(visit == "T1" | visit == "T3") %$%         
    geo_id ->
  visit_index

gsea_sub <- gsea_eset[,visit_index]

pData(gsea_sub) %>%
  mutate(visit=factor(visit, levels=c("T1", "T3"))) ->
  pData(gsea_sub)

design1 <- model.matrix(formula(form), data=pData(gsea_sub))
v1 <- voom(exprs(gsea_sub), design=design1)
btm_camera <- as_tibble(camera(v1, btm_ind, design1, 
                               contrast="visitT3"), rownames="geneset")
num_sig <- c(num_sig, sum(btm_camera$FDR <= fdr_cut))

btm_camera %>% 
  mutate(time = "T3") %>%
  rbind(all_res) ->
  all_res

btm_camera %>% 
  filter(FDR < fdr_cut) %>% 
  headerKable() %>%
  kableExtra::scroll_box(height="300px")
```
  
### T1 versus T4 (T4 - T1)  
  

```{r T1vsT4}

## get subset
pData(gsea_eset) %>% 
  filter(visit == "T1" | visit == "T4") %$%         
    geo_id ->
  visit_index

gsea_sub <- gsea_eset[,visit_index]

pData(gsea_sub) %>%
  mutate(visit=factor(visit, levels=c("T1", "T4"))) ->
  pData(gsea_sub)

design1 <- model.matrix(formula(form), data=pData(gsea_sub))
v1 <- voom(exprs(gsea_sub), design=design1)
btm_camera <- as_tibble(camera(v1, btm_ind, design1, 
                               contrast="visitT4"), rownames="geneset")
num_sig <- c(num_sig, sum(btm_camera$FDR <= fdr_cut))

btm_camera %>% 
  mutate(time = "T4") %>%
  rbind(all_res) ->
  all_res


btm_camera %>% 
  filter(FDR < fdr_cut) %>% 
  headerKable() %>%
  kableExtra::scroll_box(height="300px")
```


### T1 versus T5 (T5 - T1)  
  
There is a confound between visits T5 and the other visit times. The T5 visit used the HGU133A 2.0 chip while the other visits used the HGU133 Plus 2.0 chip. Boxplots of normalized scores and MDS plots show a clear separation between the samples for the two chips (see Quality Control section).  
  
```{r T1vsT5}

## get subset
pData(gsea_eset) %>% 
  filter(visit == "T1" | visit == "T5") %$%         
    geo_id ->
  visit_index

gsea_sub <- gsea_eset[,visit_index]

pData(gsea_sub) %>%
  mutate(visit=factor(visit, levels=c("T1", "T5"))) ->
  pData(gsea_sub)

design1 <- model.matrix(formula(form), data=pData(gsea_sub))
v1 <- voom(exprs(gsea_sub), design=design1)
btm_camera <- as_tibble(camera(v1, btm_ind, design1, 
                               contrast="visitT5"), rownames="geneset")
num_sig <- c(num_sig, sum(btm_camera$FDR <= fdr_cut))

btm_camera %>% 
  filter(FDR < fdr_cut) %>% 
  headerKable() %>%
  kableExtra::scroll_box(height="300px")
```
  
# Heatmap of significant gene-sets

```{r heat_params}
new_fdr <- 0.00001
```

FDR cutoff <= `r new_fdr`  
  
```{r heatmap}

all_res %>% 
  filter(FDR < new_fdr) %$%
  geneset ->
  sigs

all_res %>%
  filter(geneset %in% sigs) %>%
  mutate(sign = as.numeric(plyr::mapvalues(Direction, 
                                           from=c("Up", "Down"), 
                                           to=c(1, -1)))) %>%
  mutate(FDR_sign = -log10(FDR)*sign) ->
  genies
  
dcast(genies, geneset~time, value.var="FDR_sign") %>%
  as.tibble() %>%
  column_to_rownames("geneset") %>%
  as.matrix() ->
  heat_mat

heatmap3(heat_mat, Rowv=TRUE, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.65 , showRowDendro=FALSE, 
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))

genies %>%
  arrange(FDR) %>%
  select(-c(sign, FDR_sign)) %>%
  head(30) %>%
 headerKable() %>%
  kableExtra::scroll_box(height="300px")
  
```
  
  
## Summary  
  
```{r summary}
tibble(visit=sort(unique(gsea_eset$visit))[-c(1,2)], "significant genesets"=num_sig) %>%
  headerKable()

```
