---
title: Parasitemia classification with pamr
author: Carl Murie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document

---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)

library(Biobase)
library(tidyverse)
library(here)
library(RNASeqUtilities)
library(magrittr)
library(pamr)
```

```{r read}
vahey_eset <- readRDS(here("data", "gpl_eset.Rds"))
```

#Classification with PAM

PAM analysis was applied to samples from visit T5 and their parasitemia status was either Protected, NonProtected or Delayed.  
  
##Sample sizes  
  

```{r sample-size}

pData(vahey_eset) %>%
  filter(visit=="T5") %>%
  filter(parasitemia=="Protected" | 
           parasitemia=="NonProtected" | 
           parasitemia=="Delay") %$%
  geo_id ->
  subset_ids

vahey_sub <- vahey_eset[,subset_ids]

tab <- countSampleSizes(pData(vahey_sub), "visit", "parasitemia")
headerKable(tab, labels="parasitemia", levels=3)
```

##Apply PAM  
  
###Training  
  
Choose an appropriate threshold to minimize classification errors. We choose a value that minimizes the classification error of both groups. 
  
```{r pam, eval=TRUE, message=FALSE}

## create pam list
vah_dat <- list(x=exprs(vahey_sub),
                y=vahey_sub$parasitemia,
                genenames=rownames(vahey_sub))

## get initial training parameters
vah_train1 <- pamr.train(vah_dat, n.threshold=40)
```


```{r cross-validate}

## use cross-validation to find optimal threshold
vah_train2 <- pamr.cv(vah_train1, vah_dat)
vah_train2
```

```{r plot_threshold}
pamr.plotcv(vah_train2)
vah_threshold <- 1.358
```

Based on the second plot above and that the Vahey paper found a classification set of 393 genes I chose a threshold value of `r vah_threshold` which uses 401 genes.
 
##Plot similar to Vahey paper  
  
```{r pam_plot, eval=TRUE} 
pamr.plotcvprob(vah_train1, vah_dat, threshold=vah_threshold)


```

This plot is similar to the one in the Vahey paper.  
  
